---
title: 'Search Files in RStudio'
author: ~
date: '2019-11-08'
slug: funneljoin
categories: []
tags: []
description: 
draft: true
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## funneljoin

[Emily Robinson](https://hookedondata.org) has an awesome new package
[funneljoin](https://github.com/robinsones/funneljoin). I'm excited to see 
her [slides](https://twitter.com/drob/status/1192834052203667456?s=20), but 
in the meantime I dove into her
[vignette](https://robinsones.github.io/funneljoin/articles/funneljoin.html).

Her examples are around ads, but customer usage funnels are also important 
for product design. We want to understand the paths customers take to get
end results, and we want to know if bad design is causing customers
to fall off. 

I'm so excited about Emily's package, because this is the first 
systematic approach to funnel analysis I've seen. 

## Business Example

```{r}
library(dplyr)
library(funneljoin)
library(purrr)
library(tidyr)
library(lubridate)
library(ggplot2)

set.seed(24)
```

```{r, include=FALSE}
week_to_days <- list(
  "one" = 0,
  "two" = 7,
  "three" = 14,
  "four" = 21
)
```


```{r, include=FALSE}
user_weeks <- tibble(
  user = rep(1:10, 3 * 4),
  weeks = map(names(week_to_days), rep, 10 * 3) %>%
    flatten_chr(),
  days_after_start = map(flatten_dbl(week_to_days), ~rep(., 10 * 3)) %>% 
    flatten_dbl(),
)
```

```{r, include=FALSE}
upload_weekly_props <- list(
  "one" = 1,
  "two" = 0.8,
  "three" = 0.6,
  "four" = 0.4
)

step_weekly_props <- list(
  "one" = 0.4,
  "two" = 0.6,
  "three" = 0.8,
  "four" = 1
)
  

wide_funnel <- user_weeks %>%
  mutate(upload = map_dbl(weeks, ~rbinom(1, 1, upload_weekly_props[[.]]))) %>%
  mutate(submit = 
    map_dbl(weeks, ~rbinom(1, 1, step_weekly_props[[.]])) * upload
  ) %>%
  mutate(print = 
    map_dbl(weeks, ~rbinom(1, 1, step_weekly_props[[.]])) * submit
  )

wide_funnel
```

```{r, include=FALSE}
event_logs <- wide_funnel %>%
  pivot_longer(upload:print, names_to = "event") %>%
  filter(value != 0) %>%
  select(-value)

event_logs
```

```{r, include=FALSE}
logs <- event_logs %>%
  mutate(skew = rbinom(n(), 1, 0.2)) %>%
  mutate(date = case_when(
    event == "upload" ~ as.Date("2020-01-06") + skew + days_after_start,
    event == "submit" ~ as.Date("2020-01-08") + skew + days_after_start,
    event == "print" ~ as.Date("2020-01-10") + skew + days_after_start,
  )) %>%
  arrange(date) %>%
  select(date, user, event)
```

Suppose we have a logs dataset, indicating events on a platform. Our website is 
a platform where customers can upload data, submit it to be analyzed, and print 
a report. Our customer only gets value out of the platform when they print a 
report, so we want to know what percentage of customers are printing after
uploading. 

However, not every customer who uploads will print a report. 
1. system reliably could be low. They may have uploaded, but submissions are
closed.

2. turnaround time is too slow. Customers need reports to take action the
week they upload. Even a few days late would be too slow. So customers may
submit reports, but don't bother to print them because they came too late to 
be useful.

3. there may be a seasonal component. To meet quotas, printing reports may
be more important later in the month.

Therefore, we need to be careful analyzing the data. If customer's first report 
is in January, but they didn't print out a report until February, our platform
may be failing them. 

We'll show a few ways that `funneljoin` can help us with this analysis.

### Customer Logs

Here's a dataset that records each upload, submission, and printing for our 
users.

```{r}
logs
```

Let's analyze do some funnel analysis. I'll make this a function, since I'll be
reusing the same funnel through-out the blog post.

```{r}
customer_funnel <- function(.data, type, ...) {
  .data %>% 
    funnel_start(
      moment_type = "upload",
      moment = "event",
      tstamp = "date",
      user = "user"
    ) %>%
    funnel_steps(
      moment_types = c("submit", "print"),
      type = type,
      gap_col = TRUE,
      ...
    ) %>%
    mutate(days_to_submit = .gap.x / 60 / 60 / 24) %>%
    mutate(days_to_print = .gap.y / 60 / 60 / 24)
}
```

```{r}
logs %>%
  customer_funnel("first-firstafter") %>%
  summarise(mean(days_to_submit), mean(days_to_print))
```

But this is only looking at the time between the first upload, and the first
submit/print afterwards.

If we look at all uploads, they have a much shorter conversion gap. 

```{r}
logs %>%
  customer_funnel("any-firstafter") %>%
  summarise(mean(days_to_submit), mean(days_to_print))
```

What about the time to print within a 7 day window? Our customers need to
print out that report this week. 

```{r}
weekly_customer_funnel <- logs %>%
  customer_funnel(
    "any-firstafter", max_gap = as.difftime(7, units = "days")
  )

weekly_customer_funnel %>%
  summarise(mean(days_to_submit, na.rm = TRUE), mean(days_to_print, na.rm = TRUE))
```

```{r}
weekly_customer_funnel
  summarize_funnel()
```

But how do those conversions trend by week? One of our customer behaviors is that
the end of the month is much more important to them. 

```{r}
weekly_customer_funnel %>%
  mutate(upload_week = floor_date(date_upload, "week")) %>%
  group_split(upload_week) %>%
  map_dfr(summarize_conversions, date_print)
```

This confirms our suspicions that conversions is highly correlated with week.

```{r}
weekly_customer_funnel %>%
  mutate(upload_week = floor_date(date_upload, "week")) %>%
  nest(funnel = -upload_week) %>%
  mutate(funnel = map(funnel, summarize_conversions, date_print)) %>%
  unnest(funnel) %>%
  arrange(upload_week)
```

Instead of using `max_gap`, you could also nest by upload week first, build
and find the conversion for each nested data-frame. 

## What I'd like to see

As much as I love `funneljoin`, I think the 
> `funnel -> nest -> map(summarize_conversion) -> unnest`

workflow is a little clunky. 

I think it would be cool if `funnel_start` respected a grouped data frame,
when looking for the next step in the funnel.

For example, we can pass a grouped data frame to `funnel_start`:

```{r}
logs %>%
  mutate(week = week(date)) %>%
  group_by(week) %>%
  funnel_start(
    moment_type = "upload",
    moment = "event",
    tstamp = "date",
    user = "user"
  ) %>%
  funnel_steps(
    moment_types = c("submit", "print"),
    type = "any-firstafter"
  ) %>%
  select(date_upload, date_print)
```

But the output for `date_print` looks for the `firstafter` outside of the week
of the upload (the group parameter for the data-frame).

## Conclusion

I absolutely love this package. Funnel plots and dashboards are frequently
requested by data analysts, and this package provides a systematic interface
to cleaning and analyzing this type of data.

Definitely check it out the next time you need to create a funnel!