---
title: 'Beyond Reprex: Units Tests with callr'
author: ~
date: '2018-03-09'
slug: reprex-callr
categories: []
tags: []
description: 
draft: true
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## Introduction

I've been [thinking](https://github.com/tmastny/leadr/tree/master/tests/testthat) [a lot](https://github.com/rstudio/reticulate/pull/167/files#diff-9e2fec04c626523d0a05bcbf796135cf)
about unit-tests recently: I've used them in my new `leadr` package, and to debug and test new features for [`reticulate`](https://github.com/rstudio/reticulate/pull/167) and [`knitr`](https://github.com/yihui/knitr/pull/1518). 

In many ways, unit tests are the ultimate [reprex](http://reprex.tidyverse.org/articles/reprex-dos-and-donts.html) (reproducible example). Except rather than helping some one else to reproduce your problem, a unit test graciously reproduces the issue until your are able to fix the problem.  

> "Whenever you are tempted to type something into a print statement or a debugger expression, write it as a test instead. — Martin Fowler" --- Hadley Wickham, [*R Packages*](http://r-pkgs.had.co.nz/tests.html)

In this post, I'd like to share what I've learned about testing in R and why I think more [`callr`](https://github.com/r-lib/callr) could improve unit tests.

## Function vs. Process

Let's suppose we are testing knitr. The workhorse of the package is `knitr::knit`, which converts the `.Rmd` to `.md`. We can come up with various ways to test `knit`, but that's only testing the package function, not necessarily the process. 

The actual knitr workflow for most people is clicking the **Knit** button in RStudio^[I've looked for the actual **Knit** button code in the RStudio [github repo](https://github.com/rstudio/rstudio) but couldn't find anything. I found references to the **Knit** button object, but not the actual code to run it. Maybe it is proprietary?] or using a shortcut like `cmd+shift+k`. 

But if we dive deeper into knitting (alluded to in this [github issue](https://github.com/rstudio/rmarkdown/issues/17#issuecomment-37241816) and in these [slides](https://slides.yihui.name/2017-rstudio-conf-rmarkdown-Yihui-Xie.html#6)), we see that the **Knit** button is not the `knit` function. Most importantly (for our purposes, we'll ignore `rmarkdown::render`), the **Knit** button starts a new R session. 

To come up with an effective unit test, we need to make sure we can replicating the process, not just the functional test. In this case, that means we should have one test that verifies `knitr::knit` in a new R session. 

Often the function and process are the same, but occasionally they can conflict and you'll need tests to handle that.


## Package Globals

Some packages have global side-effects. My new package [`leadr`](https://github.com/tmastny/leadr) creates new files, modifies directories, and remembers the last model you saved. [`reticulate`](https://github.com/rstudio/reticulate) starts a Python session that persists in current R environment.

When developing these packages like these, we need to generate and tests these side-effects within a unit-testing framework like [`testthat`](https://github.com/r-lib/testthat). [`callr`](https://github.com/r-lib/callr) provides a simple way to start a new R session within an R session, which allows package side-effects to be effectively tested.

```{r eval=FALSE}
callr::r(
  function() {
    rmarkdown::render(here::here(), quiet = TRUE, envir = new.env())
  })
```


The original issue, as described in [#1505](https://github.com/yihui/knitr/issues/1505), was that modified cached Python chunks did not save the Python environment between sessions. This only occured when using the **Knit** button in RStudio, or by restarting R and calling `knitr::knit` or `rmarkdown::render`. 

The reason I needed the **Knit** button or a restart to replicate the issue was because `reticulate` used package-level globals to create a Python session. So within an R session, the Python environment persisted and thus the chunk states' persisted. However, an R restart cleared the package globals and made a new Python session that didn't have the chunk's state.

This debugging lead me to [`reprex`](https://github.com/tidyverse/reprex) and [`callr`](https://github.com/r-lib/callr), cumulating in pull requests [#1518](https://github.com/yihui/knitr/pull/1518) to `knitr` and [#167](https://github.com/rstudio/reticulate/pull/167) to `reticulate`.

## Vignette Driven Development

Vignettes, as executable package code, provide another opportunity for less formal testing: error free vignette building. I've found this useful in a few contexts.

First, vignettes can be a nice way separate functions from process. 

Second, when developing a package, vignettes can be an excellent way to try out various APIs and workflows. For example, [`leadr::to_list`](https://github.com/tmastny/leadr/blob/96b83278fa45874813277ea23312f80d76b341e3/R/model_tools.R#L100) came about while writing a vignette, and I realized I needed an interface to `readRDS` to simplify reading in previously saved models. 




