---
title: Embed Slides in your blog
author: ~
date: '2018-02-02'
slug: embed-slides-knitr-blogdown
categories: []
tags: []
description: How embed slides into your blogdown website with knitr 
draft: true
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## Introduction

This blog post will show you how to embedded your slides on your blogdown website. [Yihui Xie's](https://yihui.name/) knitr makes this very:
```{r, eval=FALSE}
knitr::include_url('..')
```


## Upload your slideshow to your website

Actually, your slides just need to be uploaded somewhere, but your website is the perfect place to do it. Yihui has a has a tutorial in the [blogdown bookdown](https://bookdown.org/yihui/blogdown/static-files.html) about how to include and build your slideshows.

I recommend you also read his [slideshow tutorial](https://blogdown-static.yihui.name/slides/xaringan#1), but here is a quick summary:

1. Install [xaringan](https://github.com/yihui/xaringan), Yihui's R package for creating slideshows with R markdown.

2. Make a new slideshow with 
```{r, eval=FALSE}
File -> New File -> R Markdown -> From Template -> Ninja Presentation
```

3. Save it in your static directory.

```{r, eval=FALSE}
static/slides/First_Presentation.html
```
This will give your slideshow the following url:
```{r, eval=FALSE}
https://timmastny.rbind.io/slides/first_presentation#1
```

4. Assuming you are using the `blogdown::serve_site()` [workflow](https://bookdown.org/yihui/blogdown/rstudio-ide.html), you only need to change a few things to build your slides. First, create a new folder `R` at the top level of your website's directory that contains a script `build.R`:
```{r, eval=FALSE}
R/build.R
```
Then `build.R` only needs one line:
```{r, eval=FALSE}
blogdown::build_dir("static")
```

Now you can write, build, and maintain your slideshows on your website!

## Embed in one line of code

Embedding the slideshow is the easy part. After your slideshow is built and published by Netlify (or your equivalent service), you can embedded the slideshow with it's URL by calling a single line of code:

```{r}
knitr::include_url('https://timmastny.rbind.io/slides/first_presentation#1')
```

This includes the entire slideshow presentation as an HTML iframe in your website. This means you can treat the slides as if they were in their own window. This includes:

- copying text and images
- advancing slides


## Future work

## Acknowledgements






