---
title: Mean Differences for Meta-Analyzes
author: ~
date: '2018-02-03'
slug: meta-analysis-examples
categories: []
tags: []
description: Introduction to meta-analyzes by example
draft: false
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=TRUE, autodep=TRUE)
```

## Introduction

This post will be a short introdction to mean differences, motivated by simulated and real data.

## Mean Differences

Imagine we have a set of studies:

```{r}
library(tidyverse)
d <- tibble(study = 1:100)
```

and each study is measuring a treatment's effect on some measurement. Before the treatment, suppose we have the following measurement in each study:

```{r}
d$pre <- rnorm(100, 100, 10)
```

Now, the treatment could have a variety of effects on the measurement. Here's a few difference onces we'll look into:

```{r}
noise <- rnorm(100, 0, 2)
d$post1 <- d$pre + 10 + noise
d$post2 <- 1.15 * d$pre + noise
d$post3 <- 0.5 * d$pre + 50 + noise
```

```{r}
d %>%
  gather(post_type, post, post1:post3) %>%
  ggplot() + 
  geom_point(aes(x = pre, y = post)) +
  geom_abline(intercept = 0, slope = 1) +
  facet_wrap(~post_type)
```






