---
title: Women and the Bench Press
author: ~
date: '2018-02-01'
slug: women-weak-bench-press
categories: []
tags: []
description: How to use Git to improve blogdown and project workflow 
draft: true
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## Introduction

For raw meet data, see this website: http://www.openpowerlifting.org/ and their Github: https://github.com/sstangl/openpowerlifting


Ideas:

- Greg Nuckols looked at relative differences between men and women. Men tend to squat and deadlift roughly 50-60% more than woman, but bench 114% more. Why do men bench disproportionally more than women, compared to the other competitive lifts?
- Could it have something to do with their absolute amount of lean bodymass?
    - https://www.strongerbyscience.com/how-to-get-strong-what-is-strong/
- Greg's article on allometric scaling and the Nuckols Index
    - https://www.strongerbyscience.com/whos-the-most-impressive-powerlifter/
    
- Bench press strength related to upper-body thickness
    - https://www.strongerbyscience.com/how-to-bench/#At_the_end_of_the_day8230

I have a really good idea I need to find some time to work on. First, I want to fit a model (allometric) to the men's data. Then I input the women's weights (controlling for their body %) and use graphical posterior predictive checks to generate what the model thinks women should be capable of, and compare it to what they actually are capable of.

## Embedded Test:

here is how to embedded:

```{r}
#knitr::include_url('http://127.0.0.1:4321/slides/First_Presentation.html#1')
```





