---
title: Make a Book of recipes, feat. purrr and rsample
author: ~
date: '2018-02-20'
slug: book-recipes-rsample-purrr
categories: []
tags: []
description: Combining recipes, rsample, and purrr 
draft: true
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## Introduction 

[Max Kuhn](http://appliedpredictivemodeling.com/), the creator of the package [caret](https://github.com/topepo/caret) and author of Applied Predictive Modeling has created some cool new packages since his move to RStudio. I've been really intrigued by [rsample](https://github.com/topepo/rsample) and [recipes](https://github.com/topepo/recipes), which seem to be an movement towards a grammar of modeling.


In this post, I'd like to briefly share how we can create a "book" of recipes with tibble and purrr to prepare the recipes for cross-validation in the rsample package.

## Recipes

The purprose of the recipes package is to transform raw data frames into a matrix format ready for consumption by statistical algorithms. The package standardizations the workflow for selecting outcomes, predictors, and data preprocessing.

Max has an excellent tutorial on using recipes and rsample, but I'd like to extend the idea a litte bit. 

Instead of starting with a 

We want to create a book of recipes to evaluate different types of steps. This way we can understand if the preprocessing or variable selection effects the prediction.



```{r}
library(AmesHousing)
library(recipes)
ames <- make_ames()
```
```{r}
rec <- recipe(Sale_Price ~ Neighborhood + House_Style + Year_Sold + Lot_Area, 
              data = ames)
rec
```

```{r}
rec_training_set <- prep(rec, training = ames, retain = TRUE, verbose = TRUE)
rec_training_set
```

```{r}
bake(rec_training_set, newdata = head(ames))
```




