---
title: 'Funnel Charts in R with funneljoin - Gaps and Deadlines'
author: ~
date: '2019-11-09'
slug: funneljoin-gaps-deadlines
categories: []
tags: []
description: 
draft: true
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## funneljoin

In my previous [blog post](/blog/funnel-join), 
I explored the awesome [funneljoin](https://github.com/robinsones/funneljoin)
package. 
Check that post out for an overview of the package
and some examples.

In this post, 
I'm going to investigate business problems that can't be modeled with
gaps,
and how we can work around that.

## Gaps vs. deadlines

First, I'm going to try to come up with definitions for gap and deadline.

Gap

* A gap is a period of time that the next event in the funnel is relevant.

For example,
if a customer clicks on an ad, 
and then purchases the product 30 days later
the company may not want to attribute that purchase to the ad.

Deadline

* A deadline is a date that the next event in the funnel
  must happen before to be relevant.

For example,
Turbo Tax wants to know the conversion rate between uploading your
W-2s and filing your taxes. 
But they only care if you file your taxes before April 15.

Another example in my line of work:
football coaches need to print off opponent reports before their upcoming game.

## How to model deadlines

Gaps and deadlines model different business funnels,
but out of the box `funneljoin` only supports gaps. 

Let's figure out how we can model deadlines using funneljoin.

I'm going to start with the upload/submit/report logs again:

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(funneljoin)
library(lubridate)

logs <- readRDS("funneljoin-logs.RDS")
logs
```

We want to separate the logs based on their deadline, 
which is the week of the upload. 

```{r}
nested_events <- logs %>%
  mutate(deadline = floor_date(date, "week")) %>%
  nest(events = -deadline)

nested_events
```

This gives us a set of events for each deadline period.

Now, we want to apply the funnel and summary for the events 
in each deadline period.

```{r}
customer_funnel <- function(.data) {
  .data %>% 
    funnel_start(
      moment_type = "upload",
      moment = "event",
      tstamp = "date",
      user = "user"
    ) %>%
    funnel_steps(
      moment_types = c("submit", "print"),
      type = "any-firstafter",
    )
}
```

```{r}
nested_events %>%
  mutate(conversions = map(
    events, 
    ~customer_funnel(.) %>%
      summarize_conversions(date_print)
  )) %>%
  unnest(conversions) %>%
  select(-events)
```

In this case, 
conversion percentage is the same whether we use gaps or deadline.

However, this isn't real data. 
The simulated data I created doesn't cover every scenario.
With real data, we need to carefully think if we want gaps
or deadlines.

## What I'd like to see

As much as I love `funneljoin`, I think the 

> `nest -> funnel -> summarise -> unnest`

workflow is a little clunky. 

I think it would be cool if `funnel_start` respected a grouped data frame,
when looking for the next step in the funnel.

For example, we can pass a grouped data frame to `funnel_start`:

```{r}
logs %>%
  mutate(week = week(date)) %>%
  group_by(week) %>%
  funnel_start(
    moment_type = "upload",
    moment = "event",
    tstamp = "date",
    user = "user"
  ) %>%
  funnel_steps(
    moment_types = c("submit", "print"),
    type = "any-firstafter"
  ) %>%
  select(date_upload, date_print)
```

But the `funneljoin` finds the `firstafter` print data outside of the week
of the upload (the group parameter for the data-frame).

I want to find the `firstafter` inside the week of the upload 
(the group parameter).

## Conclusion

Gaps or deadlines depend on the business we are trying to model.
While gaps are very easy to use with `funneljoin`, 
we can still analyze conversions 
in a deadline-based funnel
with a little extra work.
