---
title: 'dplyr 1.0.0'
author: ~
date: '2020-04-07'
slug: dplyr-1-rowwise-across
categories: []
tags: []
description: 
output:
  blogdown::html_page:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

## Introduction

dplyr 1.0.0 is [coming soon!](https://www.tidyverse.org/blog/2020/03/dplyr-1-0-0-is-coming-soon/)
I'm really excited about this release. 
I think `rowwise` and `across` are excellent new functions that make 
common operations feel more intutive. 
You can try them out for yourself by installing the development version
from [Github](https://github.com/tidyverse/dplyr/) until there is a CRAN release.

This post is just me playing around with some of the new features.

## Stop Repeating Names

I've probably typed something like this a thousand times:

```{r}
library(ggplot2)
library(tidyr)
library(dplyr, warn.conflicts = FALSE)

mtcars <- as_tibble(mtcars)

mtcars %>%
  mutate(vs = as.factor(vs)) %>%
  mutate(am = as.factor(am))
```

Here's the general pattern: I'm applying a function to a column,
but I want the column name to **stay the same**.

It also happens in summaries, where you want to find
column means, but you just want the columns to be named the same:

```{r}
mtcars %>%
  summarise(disp = mean(disp), hp = mean(hp))
```

You are always repeating

> `column_name = func(column_name)`

`dplyr::across` is a way to get around that. In mutates and summarises,
by default `across` keeps the same column name after you apply a function:

```{r}
mtcars %>%
  mutate(across(vs, as.factor))
```

This is a nice, but small benefit. The cool thing is that you can
easily apply this to any number of columns you want using [`tidyselect`](https://r4ds.had.co.nz/transform.html#select) 
and [functions](https://dplyr.tidyverse.org/dev/articles/colwise.html#basic-usage).

```{r}
mtcars %>%
  summarise(across(c(disp, hp), mean))
```

```{r}
mtcars %>%
  summarise(across(is.numeric & -disp & -hp, mean))
```


## Summary Rows

One nice thing about Excel is summary rows. 

For example, say I have the following summarised table:

```{r}
diamonds_sum <- diamonds %>%
  mutate(across(cut, as.character)) %>%
  group_by(cut) %>%
  summarise(across(price, sum), across(carat, n_distinct), n = n())

diamonds_sum
```


In Excel, I could easily add a **total row** for each column. 
You can do it with dplyr, but it doesn't feel great.

```{r}
sum_of_sum <- diamonds_sum %>%
  summarise(across(-cut, sum))

sum_of_sum
```

```{r}
diamonds_sum %>%
  bind_rows(
    sum_of_sum %>%
      mutate(cut = "total")
  )
```

It's a somewhat common problem. Here's a 
[StackOverflow post](https://stackoverflow.com/questions/39507019/add-margin-row-totals-in-dplyr-chain)
that documents various solutions. 

Funny enough, you can actually do this in the new dplyr using `rowwise`.
The intution here is to tranpose the dataframe and think of `price`, `carat` and `n` 
as rows. 
Then we can add a `rowwise` summary, tranpose back and get our total row!

So we start by transposing the dataframe: 

```{r}
transpose_df <- function(x, col, name) {
  x %>%
    pivot_longer(-{{ col }}, names_to = name) %>%
    pivot_wider(names_from = {{ col }})
}

diamonds_sum %>%
  transpose_df(cut, "feature")
```

Then use `rowwise` to add a summary column for each feature:

```{r}
diamonds_sum %>%
  transpose_df(cut, "feature") %>%
  rowwise(feature) %>%
  mutate(total = sum(c_across()))
```

Then we can transpose again to get back to our original shape:

```{r}
diamonds_sum %>%
  transpose_df(cut, "feature") %>%
  rowwise(feature) %>%
  mutate(total = sum(c_across())) %>%
  transpose_df(feature, "cut")
```

This is also instructive to see how we can tranpose data using 
the `pivot_*` family of functions, and why it's not so straight-forward to
arbitrarily tranpose data. It's easy to lose information!

