---
title: 'nba stats'
author: ~
date: '2020-05-02'
slug: nba-stats
categories: []
tags: []
description: Using emmeans to find significant pairwise differences in categorical variables. Also how contrasts work in R.
draft: true
output:
  blogdown::html_page:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE,
                      results='show', cache=FALSE, autodep=FALSE, error = TRUE)
```

## Introduction

References
1. Expected Points Gained
- https://threesandlayups.com/2018/12/09/expected-points-gained-a-framework-for-evaluating-actions-on-a-basketball-court/
- breakdown of expected points based on action of a player
- depends on location and other context
- preminarily table of values

```{r}
library(read.so)
library(tidyverse)
library(glue)
library(janitor)

epg <- glue("
|    action   |  epg  |
|-------------|-------|
| 2p fg       |   0.9 |
| 3p fg       |   1.9 |
| missed fg   | -0.85 |
| off rebound |  0.85 |
| def rebound |  0.25 |
| turnover    |  -1.1 |
| steal       |   1.1 |
") %>%
  read_md()
```

2. Mapping shots
- https://fivethirtyeight.com/features/how-mapping-shots-in-the-nba-changed-it-forever/
- shot charts to map out expected probabilities based on location

3. Expected Possession Value
- https://nbastatsgeeks.wordpress.com/newinnovative-apbrmetrics/expected-possession-value-epv/
- based on spatial configuration of players
- advanced data collection required

4. Expected Points Gained
- https://tipstersclub.com/23/10/2018/houdini/points-gained-is-this-advanced-stat-really-good-for-nba-predictions/
- "how more points a certain player gain than the average player in the NBA"
- "predict a team performance is to add all their player’s points gained"

5. NBA_PySpark
- https://github.com/spoonertaylor/NBA_PySpark
- good data source:
  128,069 shots from 281 different players in the 2014-2015 NBA season
- probability of shot completion based on factors in data.
- Looks like the data has shot distance, but not location

```{r}
temp <- tempdir()
system(glue("kaggle datasets download -q -p {temp} dansbecker/nba-shot-logs"))
shot_logs <- read_csv(
  pipe(paste0("unzip -p ", file.path(temp, "nba-shot-logs.zip")))
)

shot_logs <- shot_logs %>%
  clean_names() %>%
  glimpse()

game_logs <- shot_logs %>%
  mutate(shot_type = case_when(
    pts_type == 3 ~ "three",
    pts_type == 2 & shot_dist <= 8 ~ "paint",
    TRUE ~ "mid"
  )) %>%
  arrange(game_id, period, game_clock) %>%
  select(
    game_id, period, game_clock, location, pts_type, pts,
    shot_result, shot_type, player_name
  ) %>%
  group_by(game_id) %>%
  mutate(
    home = cumsum(ifelse(location == "H", pts, 0)),
    away = cumsum(ifelse(location == "A", pts, 0))
  ) %>%
  ungroup() %>%
  relocate(home, away, .after = game_clock)

game_logs
```

6. Creating Shot charts
- https://thedatagame.com.au/2015/09/27/how-to-create-nba-shot-charts-in-r/
  - from 2015
- Shot Charts App
  - https://github.com/toddwschneider/ballr
  - need to look through Github and make them templates

7. Shots Data
- https://github.com/toddwschneider/nba-shots-db
- 1.5 gb

8. Player Shots Data
- https://cran.r-project.org/web/packages/ballr/ballr.pdf
- NBAPlayerPerGameStats
	- aggregated season shot percentages

```{r}
library(ballr)
example_player <- NBAPlayerPerGameStats("/players/d/davisan02.html") %>%
  as_tibble()
example_player
```

8. nbastatR
- http://asbcllc.com/nbastatR/index.html
- scrapped nba data
- http://asbcllc.com/nbastatR/reference/teams_players_stats.html

```{r}
library(nbastatR)

player_stats_ex <- teams_players_stats(seasons = 2018, types = "player",
  modes = c("PerGame", "Totals"),
  tables = c("shots", "shot locations")
)

player_stats_ex %>% glimpse()
```

```{r}
player_stats_ex %>%
  pull(dataTable) %>%
  pluck(1) %>%
  glimpse()
```
