---
title: 'Intro to Hugo: The Masterchef of Layouts'
author: ~
date: '2020-06-24'
categories: []
tags: []
nohl: true
description:
slug: intro-hugo-blogdown-chef
output:
  blogdown::html_page:
    toc: true
---

<img src="/blog/intro-hugo-blogdown-chef/hugo-chef.svg"
  align="center"
  height=400
  style="background-color: white; padding-right: 0px;"
/>

## Introduction

[Hugo Themes](https://themes.gohugo.io/) and [blogdown](https://bookdown.org/yihui/blogdown/) 
make blogging as easy as writing [markdown](https://gohugo.io/content-management/formats/#learn-markdown)
or [RMarkdown](https://rmarkdown.rstudio.com/),
but in the back of my mind I've never been totally satisfied by the defaults. 
Scrolling through the themes, I could never find one that was _just_ right.
And it wasn't just the appearance. I wanted to organize and layout my website
in a way none of the existing themes offered:

* two-column blog previews
* sections for projects, and subsections within projects
* special lists and previews for certain content

So last weekend, I decided to take matters into my own hands and customize my
[existing theme](https://github.com/yihui/hugo-xmin). 
After hours of frustration, the layout of my website started coming
together. Eventually, I was even having fun, as I realized how much control I
had over the layout and appearance of my website. 

Looking back at my experience, I realized that some of my frustration stemmed 
from the fact that I didn't really understand what Hugo _was_. 
Hugo's own answer to [What is Hugo](https://gohugo.io/about/what-is-hugo/) states

> In technical terms, Hugo takes a source directory of files and templates and uses these as input to create a complete website.

I'm sure this is technically true, but it didn't really help me understand the 
conceptual purpose of Hugo. 
So in this article, I'm going to explain what Hugo is and does by cooking metaphor, 
inspired by Alison Hill's [A Spoonful of Hugo](https://alison.rbind.io/post/2019-02-19-hugo-archetypes/)
blog series. 

Hugo is a master chef: it follows recipes to build
gourmet websites, and helps you design and build a menu to showcase all your
great content. I'll breakdown two important components of websites, pages and lists,
and explain using food metaphors how Hugo lets you define, modify, and expand
the layout options. 

## Pages (and sandwiches)

A restaurant needs food and websites need content. With Hugo and 
[Hugo Themes](https://themes.gohugo.io/), you can focus on writing the content,
while Chef Hugo makes the rest of the dish. 

While the content is the most important part of the page, there are other
pieces of information and iconography on a website. In general, there are
3 sections to a webpage. I've used an example from my own website below:

![](webpage.svg)

The body is your blog post. All the words, images, and code you want to share
with the world. It's converted into what you see from your `*.md` or `*.Rmd` 
source.

The header and footer _sandwich_ the body with other relevant information
that is the same for all other posts. 
In my case, the header is the connection to the rest of my website, with
links to my home, blog, and projects.
The footer contains things that go at the end, like comments and additional
links off-site. 

Headers and footers also serve a purpose in web-development. 
Yihui Xie has an [instructional PR](https://github.com/yihui/hugo-xmin/pull/5)
on syntax highlighting showing that styling (CSS) goes in the header, 
and JavaScript goes in the footer. If you want to style your page, or 
add JavaScript elements, you'll want to dig in there.

Chef Hugo is responsible for combining these ingredients into an actual webpage: 
![](hugo-sandwich.svg)

The recipe Chef follows is found in the theme
and is constructed with [Hugo Templates](https://gohugo.io/templates/introduction/).
Once again, Yihui Xie has another good example in this 
[snippet of code](https://github.com/yihui/hugo-xmin/blob/982dfb320e8ac58a78203be76c5b06beae25d4f9/layouts/_default/single.html#L1-L12):


```{r echo=FALSE}
html <- '{{ partial "header.html" . }}
<div class="article-meta">
<h1><span class="title">{{ .Title | markdownify }}</span></h1>
{{ with .Params.author }}<h2 class="author">{{ . }}</h2>{{ end }}
{{ if (gt .Params.date 0) }}<h2 class="date">{{ .Date.Format "2006/01/02" }}</h2>{{ end }}
</div>


<main>
{{ .Content }}
</main>


{{ partial "footer.html" . }}'

blogdown::shortcode_html(
  "highlight", "html", "style=pastie,hl_lines=1 10 14,linenos=true", .content = html
)
```

These 14 lines of code make up the hugo template [hugo-xmin](https://github.com/yihui/hugo-xmin)
uses to create a webpage. Lines 1, 10, and 14 are where the real action happens.
These lines tell Hugo to insert the body between header and footer to complete
the sandwich. 

This template is the starting point for your page layout, 
and you can modify it however you'd like! You can change how this template
combines the header and footer with page data like titles or dates, or you
can dive into a
[`partial`](https://gohugo.io/templates/partials/)
to modify the header's content or style!
It's really fun with `blogdown::serve_site`, since you can see your changes
happen almost in real-time.
Modifying your existing theme is a great way to learn about Hugo and web-design.
My own website [theme](https://github.com/tmastny/hugo-xmin)
evolved from Yihui's instructional theme.

## Lists (and menus)

As a Master Chef, Hugo knows some people want to make more than one sandwich; 
they want a whole _menu_ to build a restaurant. 

The same is true for a blog: you'll need a menu showcasing all your posts
so people can find and read them. I have two "menus" on my website. 

![](hugo-lists.svg)

Menus, posts, pages: at the end of the day they are all _lists_. 
Hugo knows you need it, so they offer 
[list templates](https://gohugo.io/templates/lists/).
These are the instructions that tell Hugo how to find all your posts (sandwiches) 
and list them all in one place (a menu).

![](hugo-menu.svg)

Diving into the [list template](https://github.com/tmastny/hugo-xmin/blob/f70eb23c496cfe5a4ab51c9f27d487641313b3bd/layouts/_default/list.html#L5-L21)
for my own theme, you can see that it's fundamentally powered by a `for` loop: 

```{r echo=FALSE}
html <- '{{ partial "header.html" . }}


{{ .Content }}


<ul>
  {{ $pages := .Pages }}
  {{ $paginator := .Paginate (where $pages "Section" "!=" "") }}
  {{ range $paginator.Pages }}
  <li>
    <span class="date">{{ .Date.Format "2006/01/02" }}</span>
    <a href="{{ .RelPermalink }}">{{ .Title | markdownify }}</a>
  </li>
  {{ end }}
</ul>
{{ template "_internal/pagination.html" . }}


{{ partial "footer.html" . }}'

blogdown::shortcode_html(
  "highlight", "html", "style=pastie,hl_lines=1 10-15 20,linenos=true", 
  .content = html
)
```

The `range` function iterates over a set of pages you choose 
and inserts the html providing the date, title, and link
to the blog post in a bullet point.

This list, like the blog post in the last section, is still a webpage, 
so it's also sandwiched by the header and footer on lines 1 and 20.

Hugo provides all the tools to create lists anyway you desire. 
You can include or exclude sections or individual posts. 
I've even extend it to create a [two-column list](https://github.com/tmastny/hugo-xmin/blob/f70eb23c496cfe5a4ab51c9f27d487641313b3bd/layouts/partials/two-column.html#L10-L30)
that you see on my homepage.

## Dinner is served

In truth, this article is just the appetizer. 
There's a lot to learn about Hugo, but making simple changes to templates
in existing themes is a great way to get started. 

Credit goes to Yihui's [instructional Hugo Theme](https://yihui.org/en/2017/11/tweak-blogdown-site/).
The theme design is straighforward, and there are breadcrumbs throughout, 
guiding you through the technical details.
And Hugo's excellent [documentation](https://gohugo.io/documentation/) provides
comprehensive coverage of all the functions and parameters, as well as many
examples to learn from and tweak.
