---
title: 'Intro to Hugo: The Masterchef Behind Easy Blogging'
author: ~
date: '2020-06-24'
categories: []
tags: []
nohl: true
description:
slug: intro-hugo-blogdown-chef
output:
  blogdown::html_page:
    toc: true
---

<img src="/blog/intro-hugo-blogdown-chef/hugo-chef.svg"
  align="center"
  height=400
  style="background-color: white; padding-right: 0px;"
/>

## Introduction

Hugo appeals to writers, data scientists, and programmers, because of the focus
on content. You don't need to know `html` or CSS to write a post. In fact,
you get to use familiar formats like [markdown](https://gohugo.io/content-management/formats/#learn-markdown)
or [RMarkdown](https://rmarkdown.rstudio.com/), to interweave your writing
with code, figures, and equations.

For web developers, Hugo offers a customizable framework to create websites
with reusable templates. Websites like [tidyverse.org](https://github.com/tidyverse/tidyverse.org)
and [ropensci.org](https://ropensci.org/technotes/2019/01/09/hugo/),
which manage blogs with many contributing authors, are able to use
templates to make it easy for authors to [submit posts](https://alison.rbind.io/post/2019-02-19-hugo-archetypes/).

There are three components to a website:

1. content
2. styling
3. layout

[Hugo Themes](https://themes.gohugo.io//theme/hugo-whisper-theme/post/)
and tools like [blogdown](https://bookdown.org/yihui/blogdown/) 
are perfect for the writer, programmer, or data scientist who wants to start
to start blogging. They let the blogger use familiar formats like [markdown](https://gohugo.io/content-management/formats/#learn-markdown)
or [RMarkdown](https://rmarkdown.rstudio.com/), to interweave writing
with code, figures, and equations. No html or CSS experience required. 

The next stage of customization is styling. This is using html and CSS to change
colors, shapes, and fonts of the website, tailoring it to your desire  







But what about the data scientist, like me, who doesn't have much experience
with the web. What does Hugo actually do, and why should I care?
Hugo's answer to [What is Hugo](https://gohugo.io/about/what-is-hugo/) states

> In technical terms, Hugo takes a source directory of files and templates and uses these as input to create a complete website.

which is all true, but doesn't really help me grasp to overall conceptual 
purpose of Hugo.

To explain Hugo in _non-technical terms_, let me introduce Master Chef Hugo!
Chef is opening the kitchen doors, and showing how it works behind the scenes.



and clarified two key objectives that Hugo offers tools to solve: 
page and list creation. 
I hope this article provides some clarity to you as well!







When I was first started to customize my website, I struggled
to understand what Hugo was at a conceptual-level. 
While Hugo has excellent technical [documentation](https://gohugo.io/documentation/),
it many ways it kinda assumes you already know what Hugo does. 
The



, especially in R data science, because you can focus
on the content, and not be distracted or overwhelmed by web-design or front-end
development. Posts are written in familiar formats like 
[markdown](https://gohugo.io/content-management/formats/#learn-markdown) or
RMarkdown, letting you easily combine text, code, and figures. 


can be written in .
Or you can use blogdown or [hugodown](https://github.com/r-lib/hugodown)
to compile RMarkdown right into a blog post, including all your code and figures.
This magic lets writers, programmers, and data scientists focus on writing
without being overwhelmed with web-design and front-end development.

The saying goes, you don't want to learn how to sausage is made, but this time
is an exception: Hugo is a _Master Chef_ who would love to invite you into the
kitchen. If you've ev

The utensils are clean and [well-documented](https://gohugo.io/documentation/),
and the recipes are fun to follow. We'll go over what Hugo does to build
a webpage, and how Hugo maintains a menu of the latest blog posts. 



Why learn about Hugo? If you’ve ever wanted to customize your website,
it will help to understand what Hugo does and how it can help you.
Ideally, you’ll feel more confident making changes,
and maybe take a stab at customizing your theme.


## Pages (and sandwiches)

A restaurant needs food and websites need content. With Hugo and 
[Hugo Themes](https://themes.gohugo.io/), you can focus on writing the content,
while Chef Hugo makes the rest of the dish. 

While the content is the most important part of the page, there are other
pieces of information and iconography on a website. In general, there are
3 sections to a webpage. I've used an example from my own website below:

![](webpage.svg)

The body is your blog post. All the words, images, and code you want to share
with the world. It's converted into what you see from your `*.md` or `*.Rmd` 
source.

The header and footer _sandwich_ the body with other relevant information. 
In my case, the header is the connection to the rest of my website, with
links to my home, blog, and projects.
The footer contains things that go at the end, like comments and additional
links off-site. 

Headers and footers also serve a purpose in web-development. 
Yihui Xie has an [instructional PR](https://github.com/yihui/hugo-xmin/pull/5)
on syntax highlighting showing that styling (CSS) goes in the header, 
and JavaScript goes in the footer.

Chef Hugo is responsible for combining these ingredients into an actual webpage: 
![](hugo-sandwich.svg)

The recipe Chef follows is found in the theme, 
and is constructed with [Hugo Templates](https://gohugo.io/templates/introduction/).
Once again, Yihui Xie has another good example in this 
[snippet of code](https://github.com/yihui/hugo-xmin/blob/982dfb320e8ac58a78203be76c5b06beae25d4f9/layouts/_default/single.html#L1-L12):


```{r echo=FALSE}
html <- '{{ partial "header.html" . }}
<div class="article-meta">
<h1><span class="title">{{ .Title | markdownify }}</span></h1>
{{ with .Params.author }}<h2 class="author">{{ . }}</h2>{{ end }}
{{ if (gt .Params.date 0) }}<h2 class="date">{{ .Date.Format "2006/01/02" }}</h2>{{ end }}
</div>


<main>
{{ .Content }}
</main>


{{ partial "footer.html" . }}'

blogdown::shortcode_html(
  "highlight", "html", "style=pastie,hl_lines=1 10 14,linenos=true", .content = html
)
```

These 14 lines of code make up the hugo template [hugo-xmin](https://github.com/yihui/hugo-xmin)
uses to create a webpage. Lines 1, 10, and 14 are where the real action happens.
These lines tell Hugo to insert the body between header and footer to complete
the sandwich. 

The cool part is Hugo doesn't tell you how to make your website. 
Starting with a template like this, you can modify it any way you'd like!
And it's really fun with `blogdown::serve_site`: you can see your changes
happen almost in real-time.
Modifying your existing theme is a great way to learn about Hugo and web-design.
My own website [theme](https://github.com/tmastny/hugo-xmin)
evolved from Yihui's instructional theme.

TODO: compare to a different `single` format to show you can vary this theme.


## Lists (and menus)

As a Master Chef, Hugo knows some people want to make more than one sandwich; 
they want a whole _menu_ to build a restaurant. 

The same is true for a blog: you'll need a menu show-casing all your posts
so people can find and read them. I have two "menus" on my website. 

![](hugo-lists.svg)

Menus, posts, pages: at the end of the day they are all _lists_. 
Hugo knows you need it, so they offer 
[list templates](https://gohugo.io/templates/lists/).
These are the instructions to find all your posts (sandwiches) and
list them all in one place (a menu).

![](hugo-menu.svg)

Diving into the [list template](https://github.com/tmastny/hugo-xmin/blob/f70eb23c496cfe5a4ab51c9f27d487641313b3bd/layouts/_default/list.html#L5-L21)
for my own theme, you can see that it's fundamentally powered by a `for` loop: 

```{r echo=FALSE}
html <- '{{ partial "header.html" . }}


{{ .Content }}


<ul>
  {{ $pages := .Pages }}
  {{ $paginator := .Paginate (where $pages "Section" "!=" "") }}
  {{ range $paginator.Pages }}
  <li>
    <span class="date">{{ .Date.Format "2006/01/02" }}</span>
    <a href="{{ .RelPermalink }}">{{ .Title | markdownify }}</a>
  </li>
  {{ end }}
</ul>
{{ template "_internal/pagination.html" . }}


{{ partial "footer.html" . }}'

blogdown::shortcode_html(
  "highlight", "html", "style=pastie,hl_lines=1 10-15 20,linenos=true", 
  .content = html
)
```

The homepage, like the blog post in the last section, is still a webpage, 
so it's sandwiched by the header and footer on lines 1 and 20. 
`range` on line 10 how the list is made. It loops through all the pages
in `$paginator.Pages` and inserts the html providing the date, title, and link
to the blog post. 

Hugo provides all the tools to create lists anyway you desire. 
You can include or exclude sections or individual posts. 
I've even extend it to create a [two-column list](https://github.com/tmastny/hugo-xmin/blob/f70eb23c496cfe5a4ab51c9f27d487641313b3bd/layouts/partials/two-column.html#L10-L30)
that you see on my homepage. 

## Dinner is served

In truth, this article is just the appetizer. 
There's a lot to learn about Hugo, but making simple changes to templates
in existing themes is a great way to get started. 

Credit goes to Yihui's [instructional Hugo Theme](https://yihui.org/en/2017/11/tweak-blogdown-site/).
His suggestion to study the theme if you want to learn more about Hugo and 
blogdown is true. He left breadcrumbs throughout the project to help you 
understand all the technical details. I used his theme for a long time, 
until I got the urge to customize it and found myself learning Hugo. 
Also Hugo's excellent [documentation](https://gohugo.io/documentation/) provides
comprehensive coverage of all the functions and parameters, as well as many
examples to learn from and tweak.
