---
title: 'Tuning and Cross-validation with tidymodels and scikit learn'
author: ~
date: '2020-06-18'
slug: tidymodels-scki-learn-tuning
categories: []
tags: []
description: 
draft: true
output:
  html_document:
    toc: true 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, 
                      results='show', cache=FALSE, autodep=FALSE)
```

```{css, echo=FALSE}
body {
  max-width: 1000px;
  margin: auto;
  padding: 1em;
  line-height: 1.5em;
}

/* header and footer areas */
.menu { padding: 0; }
.menu li { display: inline-block; }
.article-meta, .menu a {
  text-decoration: none;
  background: #eee;
  padding: 5px;
  border-radius: 5px;
}
.menu, .article-meta, footer { text-align: center; }
.title { 
  font-size: 1.1em; 
  line-height: 1.1em;
}
footer a { text-decoration: none; }
hr {quietexttext
  border-style: dashed;
  color: #ddd;
}

/* code */
pre {
  border: 1px solid #ddd;
  box-shadow: 5px 5px 5px #eee;
  background: #f8f8f8;
  padding: 1em;
  overflow-x: auto;
}
code { background: #f9f9f9; }
pre code { background: none; }

/* misc elements */
img, iframe, video { max-width: 100%; }
main { hyphens: auto; }
blockquote {
  background: #f9f9f9;
  border-left: 5px solid #ccc;
  padding: 3px 1em 3px;
}

table {
  margin: auto;
  border-top: 1px solid #666;
  border-bottom: 1px solid #666;
}
table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee }



body {
  font-family: Optima, Candara, Calibri, Arial, sans-serif;
}
code {
  font-family: "Lucida Console", Monaco, monospace;
  font-size: 85%;
}
```


## Introduction

[tidymodels](https://www.tidymodels.org/) is the new framework from
Max Kuhn, David Vaughan, and Julia Silge at RStudio. It's the successor to 
the [caret](https://topepo.github.io/caret/) package, which was heavily 
featured in Max Kuhn's book 
[Applied Predictive Modeling](http://appliedpredictivemodeling.com/).

tidymodels promises a modular, extensible design for machine learning in R.
It also has a wonderful [website](https://www.tidymodels.org/) design
to help you get started as soon as possible, with a focus on examples and
case studies walking you through the components of the platform. 

In this article, I want to translate concepts between tidymodels and 
scikit-learn, Python's standard machine learning ecosystem. I hope this will
be useful for Python Data Scientists interested in R, and R users who want to
learn a little more about Python. 

I'll cover the key machine learning workflows with scikit-learn and tidymodels
side-by-side, starting with preprocessing and feature engineering, 
model tranining and cross-validation, as well as evaluation and comparison. 

I think you'll find that while both frameworks have their own unique advantages,
there are many conceptual similarities that have helped me understand both 
frameworks better.

## Testing columns

```{css, echo=FALSE}
pre, code {white-space:pre !important; overflow-x:auto}
```


Lorem ipsum.

<div class="row">
<div class="col-md-6"><h2>R sagemaker</h2>
```{r}
#1234567890#1234567890#1234567890#1234567890#1234567890#1234567890#1234567890#12
library(tidyverse)

ggplot2::diamonds
```
</div>
<div class="col-md-6"><h2>AWS Sagemaker</h2>
```{python}
import pandas as pd
import numpy as np

input_data = 's3://sagemaker-sample-data-us-east-2/processing/census/census-income.csv'
df = pd.read_csv(input_data, nrows=1000)
df.head()
```
</div>
</div>
